---
title: "Reduced DAG and g-formula for estimating the average causal effect"
author: Richard Guo
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reduceDAG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let us take a look at the main functions with the built-in examples.

```{r setup}
library(reduceDAG)
library(dagitty)
data("g.example", "g.long")
```

Let us first look at the following example. 
```{r, fig.align='center', fig.retina=2, fig.width=4}
plot(g.example)
```
This is a causal directed acyclic graph (DAG): variable A is the treatment (exposure) and variable Y is the outcome. Further, there is no bidirected edge on this graph, which means there is no latent confounder. 

To estimate the counterfactual mean $\mathbb{E} Y(a)$, one can use the g-formula, which can be derived from the graph.
```{r}
gFormula(g.example)
```
Suppose all variables take only finitely many levels. To estimate the $\mathbb{E} Y(a)$ with maximum **efficiency** (i.e., minimal variance), one can replace every factor $P(\cdot \mid \cdot)$ in the factor by the corresponding empirical $\mathbb{P}_n(\cdot \mid \cdot)$. 

However, perhaps surprisingly, this could be an overkill --- certain variables can be thrown away without affecting the quality of the estimator in large samples. The following command lists all the variables that are **uninformative**.
```{r}
getUninformativeVariables(g.example)
```
In fact, variables {I, W4} above are called *redundant*, which do not inform the estimation because they are *orthogonal nuisance*. What is truly amazing is variable W1, whose information counter-intuitively "cancels out" by a careful analysis.

Further, we can **project out** the uninformative variables. The following command returns the **reduced DAG** such that all variables are (in general) informative for causal estimation.
```{r, fig.align='center', fig.retina=2, fig.width=4}
h <- reduceDAG(g.example, verbose = FALSE)
plot(h)
```

On this smaller graph, the counterfactual mean $\mathbb{E} Y(a)$ is identified by a shorter g-formula:
```{r}
gFormula(h)
```
If we estimate $\mathbb{E} Y(a)$ with an empirical version of the above formula, which does not depend on {I, W4, W1}, we can still achieve the maximal efficiency in large samples! The reduced DAG is all that we need concerning the effect of A on Y. It tells us that we do not have to collect data on {I, W4, W1}.

Here is another example. 
```{r, fig.align='center', fig.retina=2, fig.width=4}
plot(g.long)
f <- reduceDAG(g.long)
```
